### 2017-12-10 paper/jfp-submission/response.txt

Indeed, we discuss the limitations of random generation in section
5.8. We have added a sentence there describing how we could replace
random generation with constraint-based generation.

Reviewer 2
> 5. your approximation of functions to fun is rather strong, and it is
> unclear to me how well that works. I guess since all executions start
> from the beginning of the programs keeps everything rather concrete,
> so that you do not need to invent functions. But I am not sure, and I
> would like to see a thorough discussion. I am fine with novice
> programs being short and simple since that is quite likely to be the
> case when they resort to Nanomaly, but when you do that, it should be
> clear that you do.  What may be related to this is the idea to deal
> with Currying. What about functions like foldr of which the result is
> a type variable? I know that some static work on error diagnosis
> sometimes runs into a problem of some sort and since you work at the
> dynamic level this may not be the case for you. But do say and explain
> so.

> 2. what is the limit to the size of programs you can reasonably
> handle?  What if we have a large program, is it possible to have a
> different starting point? So does a trace have to start at the start
> of the program? I guess this is important since if you start with a
> higher order function then you need to dynamically invent a function,
> and then your approximation might well be in the way. This is
> something I am really curious about.  Now you have 5% Non-parametric
> function types. Does that change in this situation?

We discuss the approximation of functions to `fun` in the
"Non-Parametric Function Type" subsection of 5.3, it is the other major
limitation of our work (the first being overloading). We do in fact have
to invent functions sometimes (the definition of `gen` says that these
must be constant functions), and this interacts poorly with higher-order
functions like `foldr`. We have added another example to the discussion
in 5.3 that illustrates the issue.

### 2017-12-22 paper/jfp-submission/response.txt

## User Study

Reviewer 1
> I was not completely convinced by the metrics used for the statistical
> analysis; I would have found it more convincing to measure the speed
> at which an incorrect program may be fixed by a student sitting in front
> of a machine. That said, some statistical analysis is better than none,
> and I am greatful that the authors attempted some evaluation.

Reviewer 2
> 7. why did you do a user study as part of an exam, and not as an
> independent study? I don't think this is a good idea.

We agree that it would be preferable to have timing data in addition to
correctness data. We did attempt a study that would have collected
timing data, but unfortunately were not able to recruit enough
participants to draw any conclusions. We have updated the text in the
"Constuct" subsection of the "Threats to Validity" to discuss the lack of
timing data as a threat.

@@

Reviewer 2
> 2. you compare only to OCAML in your student user study, which seems
> to be to be an easy target. Why not Sherrloc or Mycroft which at least
> get the location right more often? If OCAML much more often gives the
> wrong location, the message is not likely to be helpful. Indeed, I
> would expect you to only compare the systems for those cases that both
> NANOMALY and OCAML blame the same location.

Reviewer 2
> 3. does your tool help students to learn about the type system, or is
> it simply a better crutch to help them along. Did you investigate this
> at all?

These are valid concerns, we have updated the "Threats to Validity" to
discuss them.

@@

Reviewer 3
> The discussion of inter-rater reliability is strange. Why do we need
> to know if the exams are graded similarly, except to check that UCSD
> students are fairly graded? Also, why is IRR so low, given that
> these are questions with a correct answer? Given that there are only
> 60 exams in question, having a single evaluator with a clear rubric
> would eliminate both this potential source of inconsistency as well
> as the need to discuss IRR at all.

Our use of IRR is intended to account for potential bias in the
raters. One such form of bias (conscious or not) could be bias in favor
of NanoMaLy since it is new and interesting, and the raters (even if
they are not directly connected to the project) may want to see it
succeed. Due to the graphical nature of NanoMaLy's traces it is not easy
to hide whether a given answer was assisted by OCaml or NanoMaLy from
the raters. A further form of potential bias is inconsistency in
evaluating the student answers, which were long-form, English
explanations of the error/fix. There is a surprising amount of
subjectivity involved in judging, based on a couple sentences, whether
the student truly understood the error.

For these reasons we feel that our use of IRR is justified to lend
credibility to the results.

We have added a footnote to the discussion of IRR explaining our
justification.

@@

## Investigating the "safe" programs

Reviewer 1
> page 23, regarding section "witness exists", I got puzzled. The study
> there is on 50 programs. It was found that 2% of the programs admit a
> witness. I deduce this must correspond to one single program. But
> then, the next sentence tells that "slightly over half of these
> programs involved had some particular property, while the rest had
> some other property."  There semms to be some inconcistency there.

Indeed, the wording here was confusing. The percentages in this section
are presented as a fraction of *all* programs, not just the random
sample of failing programs. We have updated the text at the start of 5.3
to clarify this.

@@

## Other language features

Reviewer 1
> The tool is restricted to a small subset of OCaml, in particular it does
> not support side-effects. While many FP courses begin by focusing on
> pure programs only, there are some intringuing type errors that students
> face only when working with effectful programs, e.g. forgetting to apply
> a function to the unit argument. I speculate that it would be /not/ so
> easy to generalize the present work to deal with side effects, because
> of the need to maintain the state and report it to the user.

While side effects would certainly make the visualization more complex,
we don't believe they present a fundamental limitation to our approach.
Your example of the student forgetting to supply the unit argument is
actually quite similar to the problem posed by curried functions. Either
the function will be consumed where it's return value was expected,
which would produce an error similar to our `wwhile` example, our it
will be discarded entirely as the call was intended solely for its side
effect. In the latter case, we could simply add a restriction that any
such discarded values must be unit (it is commonly considered a best
practice to explicitly discard the results of functions called just for
their side effects with a `void` function).

We have added a discussion of imperative features to section 6.

### 2017-12-22 paper/jfp-submission/evaluation.tex
address wes comments

We have implemented a prototype of our search procedure and trace
visualization for a purely functional subset of \ocaml\ --- with
polymorphic types and records, but no modules, objects, or polymorphic
variants --- in a tool called \nanomaly.
%
We treat explicit type signatures, \eg @(x : int)@, as
primitive operations that narrow the type of the wrapped value.
%
In our implementation we instantiated \gensym\ with a simple random
generation of values, which we will show suffices for the majority of
type errors.

@@

We then instructed four annotators (one of whom is an author, the other
three are teaching assistants at UCSD) to classify the answers as
correct or incorrect.
%
We performed an inter-rater reliability (IRR) analysis to determine the
degree to which the annotators consistently graded the exams.
%
As we had more than two annotators assigning nominal (``correct'' or
``incorrect'') ratings we used Fleiss' kappa~\cite{Fleiss1971-du} to
measure IRR.\@
%
Fleiss' kappa is measured on a scale from $1$, indicating total
agreement, to $-1$, indicating total disagreement, with $0$ indicating
random agreement.

@@

There are, of course, drawbacks to our approach. Five that stand out
are:
%
(1) coverage limits due to random generation;
%
(2) dealing with explosions in the size of generated traces;
%
%(3) the inability to handle certain instances of infinite types; and
(3) our use of a non-parametric function type;
%
(4) handling ad-hoc polymorphism; and
%
(5) traversal bias.

@@

Also discussed in \S~\ref{sec:how-safe}, our approach can only support
ad-hoc polymorphism (\eg\ type-classes in \haskell\ or polymorphic
comparison functions in \ocaml) in limited cases where we have enough
typing information at the call-site to resolve the overloading. This
issue is uncommon in \ocaml\ (we detected it in around 5\% of our
benchmarks), but it would surely be exacerbated by a language like
\haskell, which makes heavy use of overloading. We suspect that either
dynamic-symbolic execution or speculative instantiation of holes would
allow us to handle ad-hoc polymorphism, but defer a proper treatment to
future work.

@@

>>>>

We have implemented a prototype of our search procedure and trace
visualization for a purely functional subset of \ocaml\ --- with
polymorphic types and records, but no modules, objects, or polymorphic
variants --- in a tool called \nanomaly.
\footnote{\url{https://github.com/ucsd-progsys/nanomaly}}
%
We treat explicit type signatures, \eg @(x : int)@, as
primitive operations that narrow the type of the wrapped value.
%
In our implementation we instantiated \gensym\ with a simple random
generation of values, which we will show suffices for the majority of
type errors.

@@

We then instructed four annotators (one of whom is an author, the other
three are teaching assistants at UCSD) to classify the answers as
correct or incorrect.
%
We performed an inter-rater reliability (IRR) analysis to determine the
degree to which the annotators consistently graded the exams.
\footnote{Measuring IRR is an established practice to account for
  potential bias among raters. The students were asked to explain the
  errors in English; judging whether they truly understood the errors
  invovled a surprising amount of subjectivity, and is thus subject to
  rater bias.}
%
As we had more than two annotators assigning nominal (``correct'' or
``incorrect'') ratings we used Fleiss' kappa~\cite{Fleiss1971-du} to
measure IRR.\@
%
Fleiss' kappa is measured on a scale from $1$, indicating total
agreement, to $-1$, indicating total disagreement, with $0$ indicating
random agreement.

@@

There are, of course, drawbacks to our approach.
% %
% Five that stand out are:
% %
% (1) coverage limits due to random generation;
% %
% (2) dealing with explosions in the size of generated traces;
% %
% %(3) the inability to handle certain instances of infinite types; and
% (3) our use of a non-parametric function type;
% %
% (4) handling ad-hoc polymorphism; and
% %
% (5) traversal bias.
%
In the sequel we discuss a selection of drawbacks, and how we might
address them in future work.

@@

Also discussed in \S~\ref{sec:how-safe}, our approach can only support
ad-hoc polymorphism (\eg\ type-classes in \haskell\ or polymorphic
comparison functions in \ocaml) in limited cases where we have enough
typing information at the call-site to resolve the overloading. This
issue is uncommon in \ocaml\ (we detected it in around 5\% of our
benchmarks), but it would surely be exacerbated by a language like
\haskell, which overloads not only functions but also numeric literals,
as well as strings and lists if one enables the respective language
extensions.
%
We suspect that either dynamic-symbolic execution or speculative
instantiation of holes would allow us to handle ad-hoc polymorphism, but
defer a proper treatment to future work.

@@

\paragraph{Side Effects}

While our implementation does not currently support side effects, it
would not be difficult to add support.
%
The search procedure is easily extended to support mutation by
maintaining a store in the evaluation relation, the main complexity
would be in extending the trace visualization to demonstrate mutation.
%
Incorporating mutation directly into our reduction-based visualization
would be difficult, as mutation would have non-local effects on the
expressions and may be difficult for students to follow.
%
Instead, we could follow the example of
%
\textsc{Python Tutor}~\cite{GuoSIGCSE2013}
%
and provide a separate visualization of the mutable store, with references
visualized as constant pointers to changing objects.

### 2017-12-22 paper/jfp-submission/introduction.tex
address wes comments

\paragraph{New Material}
This paper is an extension of our ICFP '16 paper of the same
name~\cite{Seidel2016-ul}.
%
In this extension we have focused on expanding our experimental evaluation.
%
First, in \S~\ref{sec:how-safe} we investigate the student programs for
which we were unable to synthesize a witness. We group the failures into
five categories, give representative examples, and suggest ways to
improve our feedback in these cases. Interestingly, we find that in the
majority of these failed cases, the programs do not actually admit a
witness in our semantics.
%
Second, in \S~\ref{sec:locating} we attempt to use our witnesses to
localize type errors with a simple heuristic. We treat the stuck term as
a sink for typing constraints, and the values it contains as sources of
constraints. We can then predict that either the stuck term or one of
the terms that \emph{produced} a value it contains is likely at fault
for the error. We compare our localizations to \ocaml and two
state-of-the-art type error localization tools, and find that we are
competitive with the state of the art.
%
Finally, we have also extended \S~\ref{sec:user-study} with an analysis
of the statistical significance of our user study results.

>>>>

\paragraph{Contributions Relative to Prior Publications}
This paper extends our ICFP '16 paper of the same
name~\cite{Seidel2016-ul}, focusing on the experimental evaluation.
%
First, in \S~\ref{sec:how-safe} we investigate the student programs for
which we were unable to synthesize a witness. We group the failures into
five categories, give representative examples, and suggest ways to
improve our feedback in these cases. Interestingly, we find that in the
majority of these failed cases, the programs do not actually admit a
witness in our semantics.
%
Second, in \S~\ref{sec:locating} we attempt to use our witnesses to
localize type errors with a simple heuristic. We treat the stuck term as
a sink for typing constraints, and the values it contains as sources of
constraints. We can then predict that either the stuck term or one of
the terms that \emph{produced} a value it contains is likely at fault
for the error. We compare our localizations to \ocaml and two
state-of-the-art type error localization tools, and find that we are
competitive with the state of the art.
%
Finally, we have also extended \S~\ref{sec:user-study} with an analysis
of the statistical significance of our user study results.
